Compression of interferometric radio-astronomical data

### 2013-03-28 articles/Article-compression/compression.tex
Some work on the compression article

Correlators in modern interferometric radio observatories output data at high spectral and temporal resolutions \cite{evla-perley-2011,mwa-2013-tingay,atca-broadband-backend-2011} (TODO cite LOFAR). It is often not desirable to decrease the resolution before archiving, because resolution in both dimensions provides scientific value. For example, high temporal resolution is desirable for pulsar observations~\cite{ska-pulsars-2009}, variability studies~\cite{lofar-pulsars-and-transients-2011} and advanced calibration methods that solve for the ionospheric disturbances and instrumental variabilities \cite{kazemi-clustered-cal-2013,revisiting-me-ii}. On the other hand, high frequency resolution is necessary for widefield imaging of the sky~\cite{big-book-on-synthesis-imaging}, spectral-line studies~\cite{evla-high-resolution-lines}(TODO cite more?) and accurate removal of man-made interference~\cite{lofar-radio-environment}. This can lead to data volumes of several petabytes, which have to be supported by complex network architectures to allow acceptable transfer rates \cite{lofar-processing-arch-2005} (TODO cite more). While current interferometers consist of up to a few hundred of antenna elements \cite{evla-perley-2011,mwa-2013-tingay,atca-broadband-backend-2011} (TODO cite LOFAR), future instruments will have thousands of antenna elements that will produce data at even higher resolutions or with multiple beams, thereby increasing the data volume and associated storage costs by several orders of magnitudes \cite{apertif-2010} (TODO cite SKA, ESKAP(?)). For these reasons, this paper considers the compression of correlated interferometric data.

% This might go...
Various multi-media and scientific applications have used and enhanced a variety of different compression techniques to decrease the storage costs. For audio compression, entropy-based encoding schemes (Huffman-encoding \cite{huffman}) and prediction schemes (TODO cite Chebinov polynomials) are important building blocks. Wavelet decomposition is another important building block for image and video compression algorithms (TODO cite). However, data from radio interferometric radio observatories have a low signal-to-noise ratio (SNR), especially at high resolutions (TODO cite Thompson), and noise-like signals are inherently hard to compress without loss of information (TODO cite).

@@

\section{Methods}
We consider compressing correlator output samples that consist of complex values. For radio data, each complex component is normally stored as a 32-bit IEEE 754 single precision floating point value\footnote{See e.g. ``FITS Interferometry Data Interchange Format'', AIPS memo \#114, \S4.1;\\and ``MeasurementSet definition version 2.0'', CASA Memo \#229.}. We test two compression algorithms on recorded data from several telescopes. The first method is a loss-less encoding technique using a Huffman encoding \cite{huffman}, while the second method is a lossy encoding technique in the form of a non-linear minimized-squared-error quantization scheme (TODO cite?). Both encodings are a-priory optimized for normally distributed complex samples with zero mean and a certain standard deviation. The distribution of correlated samples will approximate such a distribution more closely when the SNR is low (TODO cite) and does not contain interfering sources (TODO cite my own).

Loss-less compression is only effective for the 8-bit exponents of the floating point values, because the mantissa and sign are uniformly distributed and therefore incompressible without loss. We avoid very large dictionary sizes by applying the Huffman encoding on the 8-bit exponents only. To decrease loss of compression rate due to the integer number of bits required to represent symbols (TODO cite), the exponents of two consecutive samples are bitwise concatinated and encoded as such.

>>>>

Correlators in modern interferometric radio observatories output data at high spectral and temporal resolutions \cite{evla-perley-2011,mwa-2013-tingay,atca-broadband-backend-2011} (TODO cite LOFAR). It is often not desirable to decrease the resolution before archiving, because resolution in both dimensions provides scientific value. For example, high temporal resolution is desirable for pulsar observations~\cite{ska-pulsars-2009}, variability studies~\cite{lofar-pulsars-and-transients-2011} and advanced calibration methods that solve for the ionospheric disturbances and instrumental variabilities \cite{kazemi-clustered-cal-2013,revisiting-me-ii}. On the other hand, high frequency resolution is necessary for widefield imaging of the sky~\cite{big-book-on-synthesis-imaging}, spectral-line studies~\cite{evla-high-resolution-lines}(TODO cite more?) and accurate removal of man-made interference~\cite{lofar-radio-environment}. This can lead to data volumes of several petabytes, which have to be supported by complex network architectures to allow acceptable transfer rates \cite{lofar-processing-arch-2005} (TODO cite more). While current interferometers consist of up to a few hundred of antenna elements \cite{evla-perley-2011,mwa-2013-tingay,atca-broadband-backend-2011} (TODO cite LOFAR), future instruments will have thousands of antenna elements that will produce data at even higher resolutions or with multiple beams, thereby increasing the data volume and associated storage costs by several orders of magnitudes \cite{apertif-2010,ska-station-config-2013} (TODO cite ESKAP(?)). For these reasons, this paper considers the compression of correlated interferometric data.

Various multi-media and scientific applications have applied a variety of different compression techniques to decrease storage or transmission costs. Linear quantization can be effective for compression of noisy astronomical image \cite{astro-image-compression-2010}. Entropy-based encoding schemes such as Huffman encoding \cite{huffman} and prediction schemes (TODO cite Chebyshev polynomials) can optimize the representation of data. Decomposition into basis functions can provide a way to remove the less important bits in the data, such as the wavelet decomposition of audio data to remove tones that are nearly inaudible \cite{wavelet-audio-compression-1993}. In radio astronomy, linear quantization of the captured antenna voltages prior to correlation is common and well understood \cite{quantization-efficiency-2007}.

@@

% TODO raw data compression is different than image compression in the way that we don't care about the error in individual pixels; the total error is important!

\section{Methods}
We consider compressing correlator output samples that consist of complex values. These data commonly have a low signal-to-noise ratio (SNR), especially at high resolutions \cite{thompson-radio-interferometry}, and noise-like signals are inherently hard to compress without loss of information \cite{shannon-entropy-definition-1949}. In radio data, each complex component is normally stored as a 32-bit IEEE 754 single precision floating point value\footnote{See e.g. ``FITS Interferometry Data Interchange Format'', AIPS memo \#114, \S4.1;\\and ``MeasurementSet definition version 2.0'', CASA Memo \#229.}. We test two compression algorithms on recorded data from several telescopes. The first method is a loss-less technique using Huffman encoding \cite{huffman}, while the second method is a lossy encoding in the form of a non-linear minimized-squared-error quantization scheme \cite{least-squares-quantization-2006}. Both encodings are a-priory optimized for normally distributed complex samples with zero mean and a certain standard deviation. The distribution of correlated samples will approximate such a distribution more closely when the SNR is low (TODO cite) and does not contain interfering sources (TODO cite my own).

Loss-less compression is only effective for the 8-bit exponents of the floating point values, because the mantissa and sign follow closely a uniform distribution and are therefore incompressible without loss \cite{shannon-entropy-definition-1949}. We avoid very large dictionary sizes by applying the Huffman encoding on the 8-bit exponents only. To decrease loss of compression rate due to the integer number of bits required to represent symbols (TODO cite), the exponents of two consecutive samples are bitwise concatinated and encoded as such.
